\newcommand{\v}[1]{\mathbf{#1}}
\newcommand{\dv}[1]{\dot{\v{#1}}}
\newcommand{\ddv}[1]{\ddot{\v{#1}}}
\newcommand{\uv}[1]{\hat{\v{#1}}}

\begin{document}
\maketitle

\section{Orbit Determination Past Problems}

\subsection{Fall 2023}

\subsubsection{Problem Statement}

Assume an object in a highly eccentric inclined orbit (HEO) around the Earth. The object is reported in its Cartesian state space with a mean $[x_1,x_2,x_3,v_1,v_2,v_3]$ and the second moment $\Sigma$ in an inertial coordinate frame.

\begin{enumerate}
    \item The coordinates are reported in J2000. How is that frame defined. What is the motivation for it? Transform into the coordinates that you would need to report to your sensor. Define your sensor location. Distinguish between the geocentric and geodetic coordinates explicitly.
    \item Write out the second moment in the correct form with units S explicitly assuming the state (mean and second moment) have been propagated through an Extended Kalman filter since the last measurement. What probability density distribution does your object have? Hypothesize concrete values, justify all your assumptions and answers quantitatively and qualitatively.
    \item You have the chance to observe the HEO object using a ground-based sensor.
    \begin{enumerate}
        \item At which anomaly of the orbit are you choosing to observe? Justify all your answers qualitatively and quantitatively and explicate any assumptions.
        \item You are determining a circular orbit with your chosen observations. Compute explicitly the error that is inherent in that orbit determination.
        \item Your orbit has significant uncertainty, how does the uncertainty manifest in your circular orbit detemination? Compute explicitly.
    \end{enumerate}
    \item Assume in a ground-based sensor you can only make measurements of the velocity components of an Earth-orbiting space object. Device an initial orbit determination method for this case. Explicate any assumptions and limitations of your method.
\end{enumerate}

\subsubsection{Solution}

\end{document}
